The Weakly Nonlinear Magnetorotational Instability in a thin-gap Taylor Couette flow

### 2016-10-04 paper/thingap_mri_paper.tex

This work examines Taylor Couette flow in the thin-gap regime, an idealization in which the radial extent of the channel is very small compared to its distance from the center of rotation, i.e. $(r_2 - r_1) \ll \frac{1}{2} (r_1 + r_2)$ where $r_1$ and $r_2$ are the radii of the inner and outer flow boundaries, respectively. The thin-gap approximation eliminates curvature terms, so the domain geometry is Cartesian rather than cylindrical. The excluded curvature terms have an explicit dependence on $r$, so they make the problem more challenging both analytically and numerically. In particular, in the wide-gap geometry (i.e. true Taylor Couette flow) the base angular velocity is a function of $r$, where in the thin-gap approximation the shear flow reduces to a linear profile.

@@

\section{B. Expansion of Nonlinear Terms}\label{app:nonlinear_terms}
\beq
\mathbf{N} = \epsilon^2\mathbf{N_2} \, + \, \epsilon^3\mathbf{N_3} \, + \, \mathcal{O}(\epsilon^4)
\eeq

>>>>

This work examines Taylor-Couette flow in the thin-gap regime, an idealization in which the radial extent of the channel is very small compared to its distance from the center of rotation, i.e. $(r_2 - r_1) \ll \frac{1}{2} (r_1 + r_2)$ where $r_1$ and $r_2$ are the radii of the inner and outer flow boundaries, respectively. The thin-gap approximation eliminates curvature terms, so the domain geometry is Cartesian rather than cylindrical. The excluded curvature terms have an explicit dependence on $r$, so they make the problem more challenging both analytically and numerically. In particular, in the wide-gap geometry (i.e. true Taylor-Couette flow) the base angular velocity is a function of $r$, where in the thin-gap approximation the shear flow reduces to a linear profile.

@@

Note that, following the notation of \citet{Umurhan:2007hs}, the subscripts refer to $\epsilon$ order, $z$ order, successively, such that $\mathbf{N}_{22}$ is the second-order nonlinear term which corresponds to $e^{2 i k_c z}$ z-dependence.

Equation \ref{eq:ordere2} is solved as three separate systems of equations, one for each possible $z$ resonance:

\begin{align}
\mathcal{L}\mathbf{V}_{20} & = \mathbf{N}_{20}\\
\mathcal{L}\mathbf{V}_{21} & = - \widetilde{\mathcal{L}}_1 \partial_Z \mathbf{V}_{11} \\
\mathcal{L}\mathbf{V}_{22} & = \mathbf{N}_{22}
\end{align}

Finally, at $\mathcal{O}(\epsilon^3)$ we eliminate secular terms to close the system. Secular terms are terms which are resonant with the solution to the homogenous linear equation (Equation \ref{eq:ordere}), and which cause the higher-order solutions to grow without bound. The solvability criterion we enforce to eliminate these terms is the vanishing of the inner product of the solution to the adjoint linear homogenous equation $\mathcal{L}^\dagger \mathbf{V}^\dagger = 0$ with the nonhomogenous terms in Equation \ref{eq:ordere3}, namely

\beq\label{eq:innerprod_gle}
\langle \mathbb{V}^\dagger | \mathcal{D} \mathbb{V}_{11} \rangle \partial_T \alpha + \langle  \mathbb{V}^\dagger | \widetilde{\mathcal{G}} \mathbb{V}_{11} \rangle \alpha +  \langle \mathbb{V}^\dagger | \widetilde{\mathcal{L}}_1 \mathbb{V}_{21} + \widetilde{\mathcal{L}}_2 \mathbb{V}_{11} \rangle \partial_Z^2 \alpha = \langle \mathbb{V}^\dagger | \mathbf{N}_{31} \rangle \alpha |\alpha|^2.
\eeq

This solvability criterion derives from a corollary to the Fredholm Alternative \citep[see][for a formal definition]{Clark:2016}.

Equation \ref{eq:innerprod_gle} can be rewritten as Equation \ref{eq:gle}, the Ginzburg-Landau Equation, where the coefficients are

\beq
b = \langle  \mathbb{V}^\dagger | \widetilde{\mathcal{G}} \mathbb{V}_{11} \rangle / \langle \mathbb{V}^\dagger | \mathcal{D} \mathbb{V}_{11} \rangle,
\eeq

\beq
h = \langle \mathbb{V}^\dagger | \widetilde{\mathcal{L}}_1 \mathbb{V}_{21} + \widetilde{\mathcal{L}}_2 \mathbb{V}_{11} \rangle / \langle \mathbb{V}^\dagger | \mathcal{D} \mathbb{V}_{11} \rangle,
\eeq

\beq
c = \langle \mathbb{V}^\dagger | \mathbf{N}_{31} \rangle / \langle \mathbb{V}^\dagger | \mathcal{D} \mathbb{V}_{11} \rangle. 
\eeq

We define the adjoint operator $\mathcal{L}^\dagger$ and solution $\mathbf{V}^\dagger$ as

\beq
\langle \mathbf{V^\dagger} | \mathcal{L} \mathbf{V} \rangle = \langle \mathcal{L}^\dagger \mathbf{V}^\dagger | \mathbf{V} \rangle,
\eeq

where the inner product is defined as

\beq\label{eq:inner_product_def}
\langle \mathbf{V^\dagger} | \mathcal{L} \mathbf{V} \rangle = \frac{k_c}{2\pi} \int_{-\pi/k_c}^{\pi/k_c} \int_{x_1}^{x_2} \mathbf{V}^{\dagger*} \cdot \mathcal{L} \mathbf{V} \, \mathrm{d} x \mathrm{d} z.
\eeq

The solution to the adjoint homogenous equation has the form

\beq
\mathbf{V^\dagger} = \mathbb{V}^\dagger(x) e^{i k_c z} + c.c.
\eeq

\section{B. Expansion of Nonlinear Terms}\label{app:nonlinear_terms}

At each order in our perturbation series, lower-order MRI modes nonlinearly interact. Thus there is a nonlinear term contribution at $\mathcal{O}(\epsilon^2)$ and $\mathcal{O}(\epsilon^3)$. Here we detail the form of these nonlinear terms.

The overall nonlinear contribution to our system, written as a vector $\mathbf{N}$ in Equation \ref{eq:unperturbed_matrix_equations}, is

\beq
\mathbf{N} = \epsilon^2\mathbf{N_2} \, + \, \epsilon^3\mathbf{N_3} \, + \, \mathcal{O}(\epsilon^4)
\eeq
